---
title: "openeddy_tutorials"
author: "Ladislav Sigut"
email: "sigut.l@czechglobe.cz"
output: html_notebook
editor_options: 
  chunk_output_type: console
---

R Notebooks in RStudio can be viewed and worked with either in its raw
form (*Source*) or its interpreted form resembling web page (*Visual*).
Use the toggle above to switch to a preferred variant (*Ctrl+Shift+F4*).

Use Outline to quickly navigate through main sections (*Ctrl+Shift+O*).

Every code chunk (grey box) with R code can be run altogether using the
green arrow (play button at the top right inside the box) or line by
line using *Ctrl+Enter* (preferred).

Few additional practical tips for your work. RStudio simplifies your
life with contextual autofills using *Tab*. When you start writing
function (or object) name, you need to write just three letters included
in the name, navigate by *Up*/*Down* keys to desired name and confirm by
*Tab*. This works also for objects with internal structure when
subsetting its elements by `$`. E.g. `data$` -\> *Up*/*Down* key -\>
*Tab* allows you to select column for subsetting. This can be very
useful with large number of internal objects (plots produced by
`despikeLF`) or when constructing path to your local directory. Paths
can be either relative `"."` to your working directory (try to copy
`"."` to console, place cursor behind the dot and press *Tab*) or full
paths `"c://Program Files/"` (place cursor behind the last slash and
press *Tab*).

When you write R script (*Ctrl+Shift+N*), RStudio also allows you to
write simultaneously to multiple lines if you press *Alt* and drag to
the desired amount of lines by mouse. This can be useful for formatting
tibbles with `tribble()`.

### Read and write data with units

This chapter demonstrates common operations with data using *openeddy*.
After reading the example data set you will convert its timestamp from
character representation to date-time information recognized by R,
perform subsetting that conserves additional attributes and save data
frame to CSV in its original form.

First, let´s load *openeddy* R package.

```{r}
if (!require("openeddy")) {
  if (!require("devtools")) install.packages("devtools")
  devtools::install_github("lsigut/openeddy")
}

library(openeddy)
```

Set working directory to the folder where this document is saved.

```{r}
setwd(dirname(rstudioapi::getActiveDocumentContext()$path))
```

Read the example data set that includes units on the second row. You may
want to inspect the file in your favorite text editor to see its
structure. Also, familiarize yourself with the **function
documentation**. All *openeddy* functions have detailed documentation
with description of their purpose, intended usage and often include
**examples** where you can check your assumptions about how the function
works.

```{r}
?read_eddy
data <- read_eddy("KrP16.csv")
```

Inspect whether we ingested the data set correctly. Notice that each
column has attached attributes *varnames* and *units* with respective
values. At this point *varnames* equal original column names.

```{r}
str(data)
head(data)
```

You can notice that at this point, column timestamp is represented as a
character vector. Thus R cannot work with it as with date-time
information.

```{r}
str(data$timestamp)

# or more explicitly
class(data$timestamp)
```

The conversion to "POSIXt" class family can be achieved in base R using
`strptime()`. In the context of eddy covariance data and requirements of
*openeddy* and *REddyProc* packages we also need to validate further
properties of timestamp, implemented in `strptime_eddy()`. Timestamp has
to reported in regular intervals (typically half-hourly) and without
gaps (timestamp has to be logged even if no measurements exist, i.e with
`NA` values in whole row). For easier processing, especially plotting,
timestamp should be shifted to represent the middle of the averaging
period. Typically measurements are assigned to the end of the averaging
period (30 mins), thus must be moved backwards by 15 mins (-900 s).

```{r}
?strptime_eddy
data$timestamp <- strptime_eddy(data$timestamp, shift.by = -900)
```

You can see that R now understands the timestamp as date-time
information.

```{r}
class(data$timestamp)
head(data$timestamp)
```

R is unfortunately stripping *varnames* and *units* attributes during
certain operations, most importantly during subsetting across rows.

```{r}
# attributes are conserved when subsetting across columns
str(data[, 1:3])

# attributes are stripped when subsetting across rows
str(data[1:3, ])
```

To overcome this behavior, use `ex()` when subsetting data frames with
these attributes.

```{r}
?ex
str(ex(data, 1:3, 1:6))
```

Finally, to conclude this part of the tutorial, we will save the data
frame with units to a CSV file. To obtain file identical to the input we
need to shift the timestamp back and convert it to character class.

```{r}
data$timestamp <- data$timestamp + 900
data$timestamp <- format(data$timestamp, format = "%Y-%m-%d %H:%M", tz = "GMT")
```

Save the file and compare its size and structure with the input file
*KrP16.csv* (they should be identical).

```{r}
?write_eddy
write_eddy(data, "KrP16_resaved.csv")
```

### Remapping variable names

This chapter focuses on the reproducible renaming of variables to
conform to alternative naming strategies (conventions) of different
software or flux tower networks using `remap_vars()`. The renaming can
be performed either literally (exact matching) or based on a character
string pattern (using regular expressions). In the case of pattern
matching, multiple variables can be the source of a new variable - the
source variables will be averaged. The best is to keep the remapping in
a tabular form to easily identify the source variables and their
respective new names. In case that quality control flags are available
and differ from source variables by consistent prefix and/or suffix,
they can also be included in the output and renamed accordingly.

```{r}
# Prepare the session in case the previous chapter was skipped
if (!require("openeddy")) {
  if (!require("devtools")) install.packages("devtools")
  devtools::install_github("lsigut/openeddy")
}
if (!require("tibble")) install.packages("tibble")

library(openeddy)
library(tibble) # tibbles are improved data frames

setwd(dirname(rstudioapi::getActiveDocumentContext()$path))

data <- read_eddy("KrP16.csv")
names(data)
```

Define the pairing between the source and the new variable name. In this
case, exact match between source variable name (`orig_varname`) and
column name in `data` must be found. Otherwise new column with `NA`s is
initiated (see `"missing_var"` - `"PAR_IN_1_1_1"` pair).

```{r}
# Variable mapping definition (literal)
literal_remapping <- tribble(
  ~orig_varname,     ~renamed_varname,
  "timestamp",       "TIMESTAMP",
  "Tair",            "TA_1_1_1",
  "RH",              "RH_1_1_1",
  "GR",              "SW_IN_1_1_1",
  "missing_var",     "PAR_IN_1_1_1",
  "NEE",             "NEE_1_1_1",
)
```

Notice that `varnames` attribute still carries the original name before
remapping.

```{r}
?remap_vars # for remap_vars options and description
rmap1 <- remap_vars(data, 
                    literal_remapping$renamed_varname,
                    literal_remapping$orig_varname)
str(rmap1)
```

To keep the quality control information for your variables, you can
specify the `qc` argument.

```{r}
rmap1_qc1 <- remap_vars(data, 
                        literal_remapping$renamed_varname,
                        literal_remapping$orig_varname,
                        qc = "qc_")
str(rmap1_qc1)
```

Notice that quality control column of NEE was not captured because it
has additional suffix `"_SSITC"`. In that case the column can be added
explicitly by adapting the remapping table.

```{r}
literal_remapping <- tribble(
  ~orig_varname,     ~renamed_varname,
  "timestamp",       "TIMESTAMP",
  "Tair",            "TA_1_1_1",
  "RH",              "RH_1_1_1",
  "GR",              "SW_IN_1_1_1",
  "missing_var",     "PAR_IN_1_1_1",
  "NEE",             "NEE_1_1_1",
  "qc_NEE_SSITC",    "qc_NEE_1_1_1" 
)

rmap1_qc2 <- remap_vars(data, 
                        literal_remapping$renamed_varname,
                        literal_remapping$orig_varname,
                        qc = "qc_")
str(rmap1_qc2)
```

To demonstrate remapping using regular expressions we will generate an
example data frame `Ts_profile`. We will simulate a case of four soil
temperature (Ts) profiles in different depths (0 and 10 cm) at four
directions from the tower (N, E, S, W). Additionally, Eastern soil
temperature profile has additional temperature measurement at 20 cm
depth and the table contains also air temperature measurement from 2 m
height. The columns are ordered randomly to demonstrate that remapping
does not depend on column ordering.

```{r}
Ts_profile <- data[1:5, 1, drop = FALSE]
head(Ts_profile)
set.seed(42) # makes random numbers reproducible
cm_0 <- paste0("Ts_0.00_", c("N", "E", "S", "W"))
Ts_profile[cm_0] <- data.frame(replicate(4, rnorm(5)))
head(Ts_profile)
cm_10 <- paste0("Ts_0.10_", c("N", "E", "S", "W"))
Ts_profile[cm_10] <- data.frame(replicate(4, rnorm(5, 5)))
head(Ts_profile)
Ts_profile$Ts_0.20_E <- rnorm(5, 10)
Ts_profile$Ta_2.00 <- rnorm(5, 20)
head(Ts_profile)
varnames(Ts_profile) <- c("timestamp", cm_0, cm_10, "Ts_0.20_E", "Ta_2.00")
str(Ts_profile)
Ts_profile <- Ts_profile[sample(varnames(Ts_profile))]
head(Ts_profile)
```

For the cases when we want to select multiple columns that will be
averaged, it is always a good idea to check the correct definition of
the regular expression patterns beforehand. Make sure that the pattern
does not select more variables than you expected and also captures all
variables that you require.

```{r}
?regexp # to learn about regular expressions
grep("Ts_0.00_", names(Ts_profile), value = TRUE)
grep("Ts_0.10_", names(Ts_profile), value = TRUE)
```

The `varnames` attribute stores the original variable names and allows
you to double check that remapping was performed as expected. It also
helps you to document the output.

```{r}
# Variable mapping definition (using regular expressions)
regexp_remapping <- tribble(
  ~orig_varname,     ~renamed_varname,
  "timestamp",       "TIMESTAMP",
  "Ts_0.00_",        "Tsoil_0.00_mean",
  "Ts_0.10_",        "Tsoil_0.10_mean",
  "Ts_0.20_E",       "Ts_0.20",
  "Ta_2.00",         "Ta_2.00"
)

rmap2 <- remap_vars(Ts_profile, 
                    regexp_remapping$renamed_varname,
                    regexp_remapping$orig_varname,
                    regexp = TRUE)
rmap2
str(rmap2)
cat(paste(names(rmap2), varnames(rmap2), sep = " = ", collapse = "\n"))
```

### Merging data and filling gaps in timestamp

This chapter demonstrates "vertical" and "horizontal" merging along
timestamp. By vertical merging we mean primarily extending the time
series span, i.e. the timestamp length. Horizontal merging primarily
extends the amount of columns. The cases are separated here because
timestamp has to be validated before further use (it must be regular
without gaps).

```{r}
# prepare the session in case the previous chapter was skipped
if (!require("openeddy")) {
  if (!require("devtools")) install.packages("devtools")
  devtools::install_github("lsigut/openeddy")
}

library(openeddy)

setwd(dirname(rstudioapi::getActiveDocumentContext()$path))
```

To simulate vertical merge we will subset two small chunks from original
data. Notice that they do not follow-up and differ in width (amount of
columns). To apply `merge_eddy()` we will need to convert timestamp to
POSIXt class.

```{r}
data <- read_eddy("KrP16.csv")
data$timestamp <- strptime_eddy(data$timestamp, shift.by = -900)
(chunk1 <- ex(data, 23:25, 1:4))
(chunk2 <- ex(data, 28:30, 1:6))
```

The `merge_eddy()` will take the most common time interval between
records and fill the timestamp accordingly. Notice the resulting regular
timestamp and the full set of columns with `NA`s where records were not
provided. Further, `merge_eddy()` keeps the *varnames* and *units*
attributes.

```{r}
?merge_eddy
data_merged <- merge_eddy(list(chunk1, chunk2))
data_merged
str(data_merged)
```

A special use case of `merge_eddy()` is **filling the gaps in timestamp
of a single data frame**. The presence of gaps in timestamp would be
indicated by `strptime_eddy()` error message. However, `merge_eddy()`
requires timestamp in POSIXt class to work. Thus we need to provide an
exception in `strptime_eddy` using `allow_gaps` argument.

```{r}
data <- read_eddy("KrP16.csv")
# subset reasonable amount of data for the demonstration
(chunk <- ex(data, 17:33, 1:6))
# introduce gaps in the timestamp
(gappy_chunk <- ex(chunk, -c(2:3, 6:10, 15:16)))
# convert timestamp to POSIXt
# - set allow_gaps = FALSE to view the error message
gappy_chunk$timestamp <- strptime_eddy(gappy_chunk$timestamp, allow_gaps = TRUE)
# fill the gaps in the timestamp
merge_eddy(list(gappy_chunk))
```

To simulate horizontal merge we will separately extract subsets of
meteorological and eddy covariance data. As in all cases above, both
need to include timestamp along which we will merge them. Notice that
the timestamp is not identical for both subsets.

```{r}
data <- read_eddy("KrP16.csv")
data$timestamp <- strptime_eddy(data$timestamp, shift.by = -900)
# subset reasonable amount of meteo and eddy data
(meteo <- ex(data, 17:32, c(1:2, 8, 12)))
(eddy <- ex(data, 18:33, c(1, 28, 31, 34)))
```

See that `merge_eddy()` builds on and extends the capabilities of base R
`merge()`. While `merge()` works only with two data frames,
`merge_eddy()` can combine arbitrary number of data frames, conserves
*varnames* and *units,* and adds further validations of the output.

```{r}
?merge
(m1 <- merge(meteo, eddy, all = TRUE))
# attributes are stripped
str(m1)
```

The primary intention of `merge_eddy()` is to combine provided data
chunks in chronological order and keep only rows with unique timestamp
while dropping duplicated timestamps. Thus, for horizontal merging we
will need to relax the conditions of validation using
`check_dupl = FALSE`.

```{r}
# check the outcome of check_dupl = TRUE
(m2 <- merge_eddy(list(meteo, eddy), check_dupl = FALSE))
str(m2)
```

### Plot types for viewing time series data

This chapter focuses on *openeddy* functions that simplify plotting time
series data, i.e. data that are associated with a timestamp in POSIXt
format (not implemented through `ts` class). Although the package is
clearly focused on eddy covariance data, these plotting functions can be
useful also for fast display of other types of measurements (eg. soil
chamber CO2 efflux or sap-flow measurements).

```{r}
# you might want to check the ts class
?ts

# prepare the session
if (!require("openeddy")) {
  if (!require("devtools")) install.packages("devtools")
  devtools::install_github("lsigut/openeddy")
}

library(openeddy)

setwd(dirname(rstudioapi::getActiveDocumentContext()$path))

# load and format data
data <- read_eddy("KrP16.csv")
data$timestamp <- strptime_eddy(data$timestamp, shift.by = -900)
```

Base R allows you to plot time series right away if you provide
date-time information in POSIXt format. You can notice that

1.  the formatting is not completely satisfying,

2.  they do not show units of the plotted variable,

3.  in the presence of large outliers they can be less informative.

```{r}
plot(data$timestamp, data$H)
```

This is amended in the following *openeddy* plotting functions. They are
intended to produce a large number of plots in PDF file that is quick to
render and it only suffices to change the variable `var`.

You can see that in case of `plot_hh` the units are displayed, plot has
a title and timestamp distinguishes years and its formatting can be
changed easily. The outliers can still make plots hard to read though.

```{r}
?plot_hh
plot_hh(data, var = "H", pch = 19)
# change the formatting of x-axis
plot_hh(data, var = "H", pch = 19, interval = "2 months", format = "%Y-%b")
```

To improve the readability of plots, `plot_precheck` zooms only on the
defined quantile range (on default displaying 99% of the original data)
and affects point transparency to show overlapping values better.
Otherwise it includes all features of `plot_hh` described above.

```{r}
?plot_precheck
plot_precheck(data, "H", pch = 19)
```

The above plots provide only a limited overview of the extensive data.
In order to view daily courses of measurements we need to plot over
multiple panels with reasonable time spans. `plot_eddy` supports two
temporal resolutions: monthly and weekly.

In monthly resolution, four panels with time series of selected variable
`flux` will be displayed. `flux` **can be actually any variable with
reasonable properties for this type of plotting**. Even if there is less
than 4 months of data, panels will not be expanded to keep sensible
aspect ratios. Plotting can be done

1.  without any quality control (QC) information,

2.  zooming only on the y-axis span of good data (QC flags \< 2;
    excluded data is greyed-out)

3.  showing actual flag values by color code (black = flag 0, green =
    flag 1, red = flag 2)

```{r}
?plot_eddy
# no QC
plot_eddy(data, flux = "H", qc_flag = "none", skip = "weekly")
# focus only on good quality data  
plot_eddy(data, flux = "H", qc_flag = "qc_H_SSITC", skip = "weekly")
# distinguish flag values
plot_eddy(data, flux = "H", qc_flag = "qc_H_SSITC", test = "qc_H_SSITC", 
          skip = "weekly")
```

Notice that the y-axis limits are adaptive. In the presence of outliers
the maximum span of ylim can be set to improve readability.

```{r}
plot_eddy(data, flux = "H", qc_flag = "qc_H_SSITC", test = "qc_H_SSITC", 
          skip = "weekly", ylim = c(-50, 300))
```

In weekly resolution, two sets of weekly figures are displayed in one
frame. Each consists of three panels. The top and bottom panels show
auxiliary meteorological data that are pulled from input data frame. The
middle panel shows the selected main variable and precipitation if
present. Notice that the names of **auxiliary variables have to follow
naming conventions** described in `plot_eddy` documentation for the
automated plotting to work correctly.

```{r}
plot_eddy(data, flux = "H", qc_flag = "qc_H_SSITC", test = "qc_H_SSITC", 
          skip = "monthly")
```

`plot_eddy` can further visualize GPP and Reco if present in the input
data frame and panels for auxiliary data can be further customized (see
`?plot_eddy` for examples on how to make custom setups of top and bottom
panels).

As stated above, the main variable does not need to necessarily
represent flux.

```{r}
plot_eddy(data, "h2o_mixing_ratio")
```

To further inspect the properties of the collected time series
`ggplot_stats` can be useful to characterize statistical properties of
dependence between two variables. It was originally designed to show
robust statistics of dependencies on wind direction but it can be used
also in a more general way.

```{r}
?ggplot_stats
ggplot_stats(data, "wind_dir", "wind_speed", circular = TRUE)

# simple light response curve
summer_daytime <- data$PAR > 0 & data$DOY > 150 & data$DOY < 160
ggplot_stats(ex(data, summer_daytime), "PAR", "NEE", breaks = 10)
```

### Flagging and removal of spurious measurements

This chapter demonstrates some of the the quality control (QC) tools
that can be utilized with *openeddy*. The selected QC filter will always
produce flags marking the respective flux values as good (flags 0 -- 1)
or spurious (flag 2). Since more QC aspects have to be typically
considered, flags of multiple types of filters need to be combined and
applied. To get a better impression of how a realistic QC setup for Tau,
H, LE and NEE can look like check the [QC example EC
workflow](https://github.com/lsigut/EC_workflow/blob/master/KRP16_2_QC_2023-04-14.R).

```{r}
# prepare the session
if (!require("openeddy")) {
  if (!require("devtools")) install.packages("devtools")
  devtools::install_github("lsigut/openeddy")
}
if (!require("tibble")) install.packages("tibble")

library(openeddy)
library(tibble) # tibbles are improved data frames 

setwd(dirname(rstudioapi::getActiveDocumentContext()$path))

data <- read_eddy("KrP16.csv")
data$timestamp <- strptime_eddy(data$timestamp, shift.by = -900)
```

For the demonstration purposes let´s assume that there is an obstruction
affecting fluxes, thus data from wind sector 270 -- 290° need to be
removed. For this we can use `apply_thr` with proper `flag` mode
(`"between"`) and we should specify that in this case we work with
angles.

```{r}
ggplot_stats(data, "wind_dir", "NEE")
?apply_thr
flags <- apply_thr(data$wind_dir, c(270, 290), flag = "between", angles = TRUE)
```

Now the obtained flags need to be applied to the respective flux in
order to take effect.

```{r}
?apply_QC
data$NEE_orig <- apply_QC(data$NEE, flags)
ggplot_stats(data, "wind_dir", "NEE_orig")
# notice that we can still access the non-filtered NEE
ggplot_stats(data, "wind_dir", "NEE")
```

In another case we perhaps know from the field notes that certain period
should be excluded due to maintenance or instrument failure. The extent
of the period is exaggerated here.

```{r}
?flag_periods
flags <- flag_periods(data$timestamp, 
                      as.POSIXct("2016-08-01 12:15:00", tz = "GMT"), 
                      as.POSIXct("2016-08-20 14:15:00", tz = "GMT"))
data$NEE_orig <- apply_QC(data$NEE, flags)
plot_precheck(data, "NEE_orig", pch = 19)
```

The EddyPro processing software provides the most commonly used tests of
[steady state and integral turbulence
characteristics](https://www.licor.com/env/support/EddyPro/topics/flux-quality-flags.html)
(Foken and Wichura, 1996) as a QC column **without suffix (qc\_%flux%)**
which was `correct()`ed in the example file to **"qc\_%flux%\_SSITC"**.

In addition to these standard tests EddyPro provides further statistics
extracted from the raw data. You can obtain the full set of extractable
filters with `extract_QC`, plus some additional ones, described in the
function documentation.

```{r}
?extract_QC
QC <- extract_QC(data)
```

To evaluate the flag values of each filter you can use `summary_QC`. You
can find the explanations about filter
[prefixes](https://github.com/lsigut/EC_workflow#qc-prefixes) and
[suffixes](https://github.com/lsigut/EC_workflow#qc-suffixes) at the [EC
workflow](https://github.com/lsigut/EC_workflow) github page.

```{r}
?summary_QC
summary_QC(QC, names(QC))
```

Now we will learn how to combine different QC filters into a single one
using `combn_QC`. It could be achieved with a specification of filter
names in a simple vector. However, specification of a complete QC setup
requires definition of QC scheme across all fluxes that is the best
documented in the tabular form (data frame) or even better using tibbles
(when constructed using `tribble` its definition already shows the
table). You can see the syntax below. For demonstration purposes we will
want to combine NEE QC consisting of SSITC test, test of absolute
limits, test of high frequency data spike percentage and test of CO2 and
H2O signal strength.

```{r}
filters <- data.frame(NEE = c("qc_NEE_SSITC", "qc_SAGA_abslim", 
                              "qc_SAGA_spikesHF", "qc_GA_LI7200"))
filters

# tribble() is more self-explanatory in more complicated setups 
filters <- tribble(
  ~NEE,
  "qc_NEE_SSITC",
  "qc_SAGA_abslim",
  "qc_SAGA_spikesHF", 
  "qc_GA_LI7200"
)
filters
```

To simplify the combining, the required filter columns are expected to
be part of the input data frame. Here we will store in `data` the
complete set of extracted filters. Notice the messages that inform about
automated detection of additive filters or special handling of `NA`
values.

```{r}
data[names(QC)] <- QC
?combn_QC
# here we use the above specified NEE column with filters to combine 
data$qc_NEE_prelim <- combn_QC(data, filters$NEE)
data$NEE_orig <- apply_QC(data$NEE, data$qc_NEE_prelim)
```

You can notice that the performed filtering led to an improvement when
compared to non-filtered data but outliers are still present.

```{r}
plot_hh(data, "NEE", pch = 21)
plot_hh(data, "NEE_orig", pch = 21)
```

There exists several ways how to statistically detect outliers. The one
implemented in *openeddy* (`despikeLF`) is working with
double-differenced time series, i.e. the outlier definition is based on
the comparison of the value to its neighbors. You can notice that it
includes visualization of the detection procedure to allow closer
inspection of how the function arguments affected the flagging.

```{r}
?despikeLF
# you can play with the z argument to observe the effect on detection sensitivity
desp <- 
  despikeLF(data, var = "NEE", qc_flag = "qc_NEE_prelim",
            name_out = "qc_NEE_spikesLF", var_thr = c(-100, 100), 
            plot = TRUE, z = 7)
# use the Tab button to expand the desp$plots$ in RStudio
desp$plots$`iter 1`$night$`2016-07-01 - 2016-07-13`
```

The structure of the `despikeLF` is different if you request
`plot = TRUE` or not. Here we assign the spike detection (`desp$SD`) to
our data frame, combine it again with previous filters and apply the
resulting QC to NEE.

```{r}
data$qc_NEE_spikesLF <- desp$SD
filters <- tribble(
  ~NEE,
  "qc_NEE_SSITC",
  "qc_SAGA_abslim",
  "qc_SAGA_spikesHF", 
  "qc_GA_LI7200",
  "qc_NEE_spikesLF"
)

# update our preliminary QC
data$qc_NEE_prelim <- combn_QC(data, filters$NEE)
# apply it to NEE
data$NEE_orig <- apply_QC(data$NEE, data$qc_NEE_prelim)
plot_hh(data, "NEE_orig", pch = 19)
```

We can see that the plotted NEE overview now looks much better. Ideally,
in a full QC setup, the reasons for spurious measurements should be
identified and fluxes removed using defined filters in a reproducible
way. The QC performance is the best checked using `plot_eddy`.

```{r}
plot_eddy(data, "NEE", "qc_NEE_prelim", "qc_NEE_prelim")
```

To visualize the main causes of flux exclusion you can obtain the
summary of the QC filters in a table or plot. In both cases you can
distinguish the results for each filter separately or show their
cumulative effect. Notice that the cumulative effect is not equal to the
sum of flag 2 values across all filters. This is because multiple
filters can assign flag 2 to given flux value. The exception from this
rule are additive tests (see details in `?combn_QC` documentation).

```{r}
summary_QC(data, filters$NEE)
summary_QC(data, filters$NEE, cumul = TRUE)
summary_QC(data, filters$NEE, plot = TRUE)
summary_QC(data, filters$NEE, plot = TRUE, cumul = TRUE)
```

In the case that there still remain spurious measurements for which
automated filtering is not easy to define, manual checking can be
applied. `check_manually` allows you to go through your data and
interactively select the spurious measurements that should be
additionally flagged. **This function does not require timestamp and can
be utilized generally outside of scope of eddy covariance data**. Notice
that the options available during the checking are displayed in the
console. Choose the desired option by writing respective number in
console followed by *Enter*. To achieve reproducibility, notice that you
will get the option to save your manual QC as CSV file to the defined
path (here it is your current working directory). This will help you
also if you want to rerun the QC workflow without user interaction
(`interactive = FALSE`). In that case `*manual_QC.csv` will be
automatically reloaded.

```{r}
?check_manually
man <- check_manually(data, path = ".", 
                      vars = data.frame(
                        x = "NEE",
                        y = "PAR",
                        z = "Tair"
                      ), 
                      qc_prefix = "qc_", qc_suffix = "_prelim", 
                      interactive = TRUE, "KrP16")[-1]
```

### Aggregation of gap-filled data and uncertainty evaluation

This chapter shows how to obtain a quick summary of eddy covariance and
meteorological data by aggregation to specified time intervals.
*openeddy* allows to aggregate also gappy, but preferentially gap-filled
data and obtain the uncertainty of these estimates based on the
REddyProc products. Unit conversion and sign correction is performed
where relevant.

```{r}
# load libraries and prepare data (fetched from REddyProc)
if (!require("openeddy")) {
  if (!require("devtools")) install.packages("devtools")
  devtools::install_github("lsigut/openeddy")
}
if (!require("REddyProc")) install.packages("REddyProc")
if (!require("bigleaf")) install.packages("bigleaf")

library(openeddy)
library(REddyProc)
library(bigleaf)

# notice that timestamp is reconstructed from multiple columns
DETha98 <- fConvertTimeToPosix(Example_DETha98, 'YDH', Year = 'Year',
                               Day = 'DoY', Hour = 'Hour')[-(2:4)]
EProc <- sEddyProc$new('DE-Tha', DETha98,
                       c('NEE', 'LE', 'Rg', 'Tair', 'VPD', 'Ustar'))
names(DETha98)[1] <- "timestamp"
# center timestamp (necessary for aggregations)
DETha98$timestamp <- DETha98$timestamp - 900
```

Aggregation by averaging is probably the most typical. Under the hood, R
will aggregate over all records that have identical timestamp. Since
they are all unique now we need to reduce the timestamp using format to
a desired resolution.

```{r}
?strptime # for format options
# reduced timestamp to monthly intervals
format(DETha98$timestamp, "%b-%y")[1:6]
table(format(DETha98$timestamp, "%b-%y"))
# reduced timestamp to weekly intervals
format(DETha98$timestamp, "%W-%Y")[1:6]
table(format(DETha98$timestamp, "%W-%Y"))
```

Knowing this, you can obtain e.g. monthly averages. By default, presence
of any `NA` value in the aggregation period produces `NA`. If you do not
mind the gaps in your aggregation intervals you can suppress this by
`na.rm = TRUE`.

```{r}
?agg_mean
agg_mean(DETha98, format = "%b-%y")
(means <- agg_mean(DETha98, format = "%b-%y", na.rm = TRUE))
```

While averaging is sensible in most cases, summation might not be. Also,
the outcomes of summation might be sensible only if certain unit
conversion is performed. Check the `agg_sum` documentation to see what
conversions are available. Unit conversions are automated for the most
common cases but rely on assumptions about original variable units
rather than its explicit testing. Performed unit conversions are
reported in the console. Sign correction might be useful in case of NEE
and GPP. Aggregated productivity should be represented as a positive
value for both variables. This is further clarified for NEE by renaming
it to NEP (net ecosystem productivity). Notice again the `NA` behavior.
In this case it should not be suppressed as we would clearly bias the
results.

```{r}
(zz <- agg_sum(DETha98, format = "%b-%y", agg_per = "month-1"))
openeddy::units(zz, names = TRUE)
```

Aggregation using `mean` and `sum` are the most typical cases but you
can apply arbitrary statistical function that returns single value per
interval. Here we show how to obtain minimum and maximum monthly air
temperatures. You can test that application of `range` instead would
have undesired effects.

```{r}
# Extract minimum and maximum within the intervals
# - two notations possible: a function (min) or function name ("max")
(mins <- agg_fun(DETha98, "%b-%y", min, na.rm = TRUE))
(maxes <- agg_fun(DETha98, "%b-%y", "max", na.rm = TRUE))
plot(means$Intervals, means$Tair_mean, ylim = c(-15, 35), 
     xlab = "timestamp", ylab = "Tair")
lines(mins$Intervals, mins$Tair_min)
lines(maxes$Intervals, maxes$Tair_max)
```

We saw previously that for `agg_sum` to provide reasonable estimates, we
need to supply gap-filled data. We can achieve that using *REddyProc*
package. We will actually need the gap-filling by-products to estimate
the uncertainty of the aggregated values. We can also show how can be
evapotranspiration (ET) obtained from LE by unit conversion using
*bigleaf* package.

```{r}
# Gap-fill NEE using approximate fixed uStar threshold
EProc$sMDSGapFillAfterUstar('NEE', uStarTh = 0.3, FillAll = TRUE)

# Gap-fill all other selected variables
for (i in c('LE', 'Rg', 'Tair', 'VPD')) EProc$sMDSGapFill(i, FillAll = TRUE)

# Export results and convert latent heat (LE) to evapotranspiration (ET)
# - typical ET units are mm hour-1 independent of actual measurement interval
results <- cbind(DETha98["timestamp"], EProc$sExportResults())
LE_vars <- c("LE_orig", "LE_f", "LE_fqc", "LE_fall", "LE_fsd")
ET_vars <- gsub("LE", "ET", LE_vars)
results[, ET_vars] <-
  lapply(LE_vars,
         function(x) LE.to.ET(results[, x], results$Tair_f) * 3600)
openeddy::units(results[ET_vars]) <- rep("mm hour-1", length(ET_vars))
# Overwrite ET_fqc with proper values
results$ET_fqc <- results$LE_fqc
openeddy::units(results$ET_fqc) <- "-"
```

The gap-filled ET can be plotted using `barplot_agg`. Note that the
width of the bars reflects the relative sample size in each interval
(see that first interval is not a full week).

```{r}
# aggregate to weekly resolution
weekly <- agg_sum(results, "%W_%Y", agg_per = "week-1")
barplot_agg(weekly, "ET_f_sum", "daily")
```

The uncertainty of the estimates is evaluated for both typical
aggregation cases - averaging and summation. Note that it represents
standard deviation, thus to obtain e.g. 95% confidence interval you need
to multiply it by 1.96 ([explained
here](https://en.wikipedia.org/wiki/97.5th_percentile_point)). Unit
conversions are performed accordingly also here.

```{r}
# Aggregate uncertainty derived from look-up table standard deviation (SD)
# - sign and unit conversions are demonstrated
?agg_fsd
(unc <- agg_fsd(results, "%b-%y", agg_per = "month-1"))
lapply(unc, openeddy::units, names = TRUE)
```

Similarly, you can obtain also uncertainty for Lasslop et al. (2010)
flux partitioning based on DayTime (DT) data. Reco and GPP uncertainty
evaluation is available only for this method. Reichstein et al. (2005)
Reco model uncertainty is not exported by *REddyProc* and GPP is
computed as residual (not modelled).

```{r}
# setup and run Lasslop et al. (2010) flux partitioning
EProc$sSetLocationInfo(LatDeg = 51.0, LongDeg = 13.6, TimeZoneHour = 1)
EProc$sGLFluxPartition(suffix = "uStar")

# estimate the Reco and GPP uncertainty
results <- cbind(DETha98["timestamp"], EProc$sExportResults())
(unc_DT <- agg_DT_SD(results, "%b-%y", agg_per = "month-1"))
lapply(unc_DT, openeddy::units, names = TRUE)
```

### List of additional functions included in openeddy

Previous chapters focused on the typical tasks and solutions when
working with eddy covariance measurements. It does not represent an
exhaustive list of *openeddy* functions and examples are often
simplified. The following functions did not make it to those tutorial
examples and their usage can be better understood by checking their
documentation or their application within
[EC_workflow](https://github.com/lsigut/EC_workflow) scripts.

-   `read_EddyPro`: EddyPro full output files have a specific structure.
    This function allows you to read single or multiple CSV EddyPro full
    output files at given path and merge them together.

-   `structure_eddy`: folder structure recommended for eddy covariance
    data processing and archivation. It can be used to create a list of
    relative paths and create those directories on your hard drive.

-   `correct`: corrects the variable names (especially those produced by
    EddyPro) to more R friendly variants.

-   `round_df`: rounds the columns of numeric mode type double to
    specified (default = 6) significant digits.

-   `varnames`, `units`: get and set units and varnames of each column
    in a data frame.

-   `choose_avail`: available variable names are checked against the
    full set of expected variables and missing cases are reported.

-   `interdep`: during quality control, flux interdependence due to
    corrections and conversions is important to consider. This idea was
    proposed in Mauder et al. (2013)

-   `fetch_filter`: flags all halfhours that have longer fetch distance
    (of given percentage of contribution to the flux) than the user
    defined boundary of the region of interest (ROI). See [ROI
    boundary](https://github.com/lsigut/ROI_boundary) for details.

-   `add_st`: storage computation is outside of the scope of *openeddy*.
    This function corrects respective flux values with existing storage
    estimates. EddyPro provides storage estimates computed using
    discrete (one point) approach.

-   `set_OT_input`: creates input for gap-filling and flux partitioning
    tools implemented either offline in *REddyProc* package or
    accessible online.

-   `Griebel20_budgets`, `spti_coverage`, `spti_boot`: yearly budgets
    with different consideration of space-time equity and related
    functions. See Griebel et al. (2020) for details.
